A Combinatorial Search of Parameterized Quantum Circuit Learning for Chemical Applications

### 2025-01-24 manuscript/main.tex
$(date)

One challenge of applying classical molecular representations for quantum machine learning models is mapping the classical features, often containing hundreds or thousands of features per sample, to the number of qubits used on the quantum device.
Morgan has X featues per sample

Using the \textit{sub}-Morgan fingerprints, we explore mapping these features to 5 and 16 qubits, the number of qubits used throughout this study, using SHapley Additive ExPlanation analysis (SHAP)\cite{lundberg_unified_2017} and principal component analysis (PCA).
We found that the most accurate models, using both 5 and 16 features, were those generated by PCA, as shown in Fig. \ref{fig:BSE_classical_features_R2} and \ref{fig:BSE_bse_classical_features_MAE}.

\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\centering	
		\includegraphics[width=\linewidth]{../classical_testing/BSE/classical_funcfit}
		\caption{FIX}
		\label{fig:classicalfuncfit}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{../images/BSE/classical_features_MAE.png}
		\caption{}
		\label{fig:BSE_bse_classical_features_MAE}
	\end{subfigure}
	\hfill		
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{../images/BSE/classical_features_R2.png}
		\caption{}
		\label{fig:BSE_classical_features_R2}
	\end{subfigure}
	\caption{}
	\label{fig:bse_classical_features}
\end{figure}

The next chemical dataset explored is based on the Data-Driven Coupled-Cluster (DDCC) method found in Refs. \cite{townsend_data-driven_2019,jones_chapter_2023}.
DDCC is a machine learning-based method for accelerating coupled-cluster singles and doubles (CCSD) calculations by predicting the CCSD T2-amplitudes using features generated by a lower-level method, typically Hartree-Fock (HF) and M{\o}ller-Plesset second-order perturbation theory (MP2).
We generate a database consisting X water molecules with y features for each molecule, using Psi4\cite{parrish_psi4_2017} and Psi4Numpy\cite{smith_psi4numpy_2018}.
The orignal feature set consists of X 
\begin{equation}
	t^{ab}_{ij(\text{MP2})} = \frac{\mel{ij}{}{ab}}{\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}}
	\label{eq:MP2_t2}
\end{equation}
Like with the BSE dataset, we explore how we can reduce the initial y features down to 5 and 16 features using SHAP and PCA.
For each $t^{ab}_{ij(\text{CCSD})}$, orbital energies ($\varepsilon_{i},\varepsilon_{j},\varepsilon_{a},\varepsilon_{b}$), Coulomb and exchange integrals ($J^{i}_{a},J^{j}_{b},K^{a}_{i},K^{b}_{j}$), binary feature whether two electrons are promoted to the same virtual orbital, the initial MP2 amplitudes, along with the numerator (two-electron integrals ($\mel{ij}{}{ab}$)) and denominator $\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}$.

>>>>

\begin{figure}[H]
	\centering	
	\includegraphics[width=\linewidth]{../classical_testing/BSE/classical_funcfit}
	\caption{FIX:The performance of various molecular representations is shown using different classical machine learning models. \textit{AB} (top row) and \textit{sub} (bottom row) denote how the chemical reaction is represented. Each column denotes a different classical machine learning model, where the first column, denoted as knn, refers to \textit{k}-nearest neighbors, the second column, labeled rfr, denotes random forest regression, and krr, the last column, denotes kernel ridge regression. The x-axis of each sub-figure denotes the various molecular representations that are analyzed: Coulomb matrices (CM), Molecular ACCess Systems (MACCS), persistence images (PIs), RDKit fingerprints (RDKit), and Morgan fingerprints (Morgan).The y-axis of each sub-figure shows the performance of each model, evaluated using the R$^{2}$, where the training set is denoted using light blue and the test set is dark blue.}
	\label{fig:classicalfuncfit}
\end{figure}

An additional challenge of applying classical molecular representations for quantum machine learning models is mapping the classical features, often containing hundreds or thousands of features per sample, to the number of qubits used on the quantum device.
Initially, the Morgan fingerpints have 2048 features per sample, that need to be reduced down to 5 or 16 qubits.
We choose 5 and 16 qubits for two reasons, the first is that these were the standard number of qubits on IBM quantum devices when we started the project and the second is that reducing the number of features reduces the depth of the circuits.
To reduce the feature set from 2048 to 5 or 16 features, we explore two different methods, SHapley Additive ExPlanation analysis (SHAP)\cite{lundberg_unified_2017} and principal component analysis (PCA), as implemented in scikit-learn.\cite{pedregosa_scikit-learn_2011}
Figs. \ref{fig:BSE_classical_features_R2} and \ref{fig:BSE_bse_classical_features_MAE} show the results for the reductions using SHAP and PCA for the training and test set of using 5 and 16 features.
The initial model using 2048 features has a train and test mean absolute error (MAE) of 1.91 and 4.98 kcal/mol, with train and test R$^{2}$s of 0.99 and 0.91, respectively.
When using SHAP to reduce the feature set size to 5 features, we see that the training set has an MAE of 16.08 kcal/mol and an R$^{2}$ of 0.39, while for the test set has an MAE of 15.86 kcal/mol and an R$^{2}$ of 0.42.
When the number of features is reduced to 16 features using SHAP, we see slight improvements with train and test MAEs of 10.48 and 11.08 kcal/mol with R$^{2}$s of 0.69 and 0.68, respectively.
Using PCA, we see an improvement in accuracy for both 5 and 16 features, where the training sets have MAEs of 4.09 and 3.23 kcal/mol and the test sets have MAEs of 10.17 and 8.40 kcal/mol, respectively.
The R$^{2}$s for PCA with 5 and 16 features also shows improvement over the reductions using SHAP, with R$^{2}$s of 0.95 and 0.69 for the training and test set, respectively, using 5 features and 0.97 and 0.78 for the training and test set, respectively, using 16 features.
Due to the increased performance, despite exhibiting overfitting, we choose to use Morgan fingerprints reduced using PCA for our QML models.

\begin{figure}[H]
	\centering	
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{../images/BSE/classical_features_MAE.png}
		\caption{}
		\label{fig:BSE_bse_classical_features_MAE}
	\end{subfigure}
	\hfill		
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{../images/BSE/classical_features_R2.png}
		\caption{}
		\label{fig:BSE_classical_features_R2}
	\end{subfigure}
	\caption{Feature reduction of the BSE dataset represented using }
	\label{fig:bse_classical_features}
\end{figure}

The next chemical dataset explored is based on the Data-Driven Coupled-Cluster (DDCC) method found in Refs. \cite{townsend_data-driven_2019,jones_chapter_2023}.
DDCC is a machine learning-based method for accelerating coupled-cluster singles and doubles (CCSD) calculations by predicting the CCSD T2-amplitudes using features generated by a lower-level method, typically Hartree-Fock (HF) and M{\o}ller-Plesset second-order perturbation theory (MP2).
We generate a database consisting 199 water molecules with 31 features for each molecule, using Psi4\cite{parrish_psi4_2017} and Psi4Numpy\cite{smith_psi4numpy_2018}.

\begin{equation}
	t^{ab}_{ij(\text{MP2})} = \frac{\mel{ij}{}{ab}}{\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}}
	\label{eq:MP2_t2}
\end{equation}
Like with the BSE dataset, we explore how we can reduce the initial y features down to 5 and 16 features using SHAP and PCA.
For each $t^{ab}_{ij(\text{CCSD})}$, orbital energies ($\varepsilon_{i},\varepsilon_{j},\varepsilon_{a},\varepsilon_{b}$), Coulomb and exchange integrals ($J^{i}_{a},J^{j}_{b},K^{a}_{i},K^{b}_{j}$), binary feature whether two electrons are promoted to the same virtual orbital, the initial MP2 amplitudes, along with the numerator (two-electron integrals ($\mel{ij}{}{ab}$)) and denominator $\varepsilon_{i}+\varepsilon_{j}-\varepsilon_{a}-\varepsilon_{b}$.
